---
title: "STA 440 Case 2"
author: "Annie Lott, Wuming Zhang, William Yang"
date: "October 1, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
require(gridExtra)

kellydat = read.table("kellydat.txt", header = TRUE)
kellydat$sn0 = ifelse(kellydat$sn1 + kellydat$sn2 + kellydat$sn3 + kellydat$all4, 0, 1)
```

# Exploratory Data Analysis


## Histograms of each Indicator

```{r, echo=FALSE, warning=FALSE, message=FALSE}
all = ggplot(data=kellydat, aes(nctdel)) + geom_histogram() + ggtitle("all patients")
fail_1 = ggplot(data=kellydat[kellydat$fail == 1,], aes(nctdel)) + geom_histogram() + ggtitle("fail = 1")
fail_0 = ggplot(data=kellydat[kellydat$fail == 0,], aes(nctdel)) + geom_histogram() + ggtitle("fail = 0")
male_1 = ggplot(data=kellydat[kellydat$male == 1,], aes(nctdel)) + geom_histogram() + ggtitle("male = 1")
male_0 = ggplot(data=kellydat[kellydat$male == 0,], aes(nctdel)) + geom_histogram() + ggtitle("male = 0")
hisp_1 = ggplot(data=kellydat[kellydat$hisp == 1,], aes(nctdel)) + geom_histogram() + ggtitle("hisp = 1")
hisp_0 = ggplot(data=kellydat[kellydat$hisp == 0,], aes(nctdel)) + geom_histogram() + ggtitle("hisp = 0")
sn0_1 = ggplot(data=kellydat[kellydat$sn0 == 1,], aes(nctdel)) + geom_histogram() + ggtitle("0 symptoms")
sn1_1 = ggplot(data=kellydat[kellydat$sn1 == 1,], aes(nctdel)) + geom_histogram() + ggtitle("1 symptom")
sn2_1 = ggplot(data=kellydat[kellydat$sn2 == 1,], aes(nctdel)) + geom_histogram() + ggtitle("2 symptoms")
sn3_1 = ggplot(data=kellydat[kellydat$sn3 == 1,], aes(nctdel)) + geom_histogram() + ggtitle("3 symptoms")
all4_1 = ggplot(data=kellydat[kellydat$all4 == 1,], aes(nctdel)) + geom_histogram() + ggtitle("4 symptoms")

grid.arrange(all, fail_1, fail_0, male_1, male_0, hisp_1, hisp_0, sn0_1, sn1_1, sn2_1, sn3_1, all4_1, ncol=3)
```

To get a better sense of the distributions in the data, we first plot histograms of the nctdel times for each level of all the variables. We notice some general trends here - nctdel time seems to go down with more symptoms, the overall distribution of times for all patients is right-skewed with a mean around 1.5. However, we note that there are only 9 data points for hispanic patients, and only 6 for patients with all 4 symptoms. Inference around these variables will therefore have high variance.


